
---
title: "Bayesian GLMM Part10"
author: "Murray Logan"
date: today
date-format: "DD/MM/YYYY"
format:
  html:
    ## Format
    theme: [default, ../resources/ws-style.scss]
    css: ../resources/ws_style.css
    html-math-method: mathjax
    ## Table of contents
    toc: true
    toc-float: true
    ## Numbering
    number-sections: true
    number-depth: 3
    ## Layout
    page-layout: full
    fig-caption-location: "bottom"
    fig-align: "center"
    fig-width: 4
    fig-height: 4
    fig-dpi: 72
    tbl-cap-location: top
    ## Code
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    code-line-numbers: true
    code-block-border-left: "#ccc"
    code-copy: true
    highlight-style: atom-one
    ## Execution
    execute:
      echo: true
      cache: true
    ## Rendering
    embed-resources: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r}
#| label: setup
#| include: false
#| cache: false

knitr::opts_chunk$set(cache.lazy = FALSE,
                      tidy = "styler")
options(tinytex.engine = "xelatex")
```


# Preparations

Load the necessary libraries

```{r}
#| label: libraries
#| output: false
#| eval: true
#| warning: false
#| message: false
#| cache: false
library(broom)     #for tidy output
library(knitr)     #for kable
#library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(tidyverse) #for data wrangling
library(brms)
library(tidybayes)
library(bayesplot)
library(broom.mixed)
library(rstan)
library(patchwork)
library(DHARMa)
source('helperFunctions.R')
```


# Scenario

For over 35 years, the AIMS long term monitoring program has performed
benthic surveys of coral reefs on the Great Barrier Reef (GBR). To do
so the team uses two main survey techniques 1) Manta Tow and 2) Photo
transects. The current example focuses on data collected using the
later technique.
Within each reef, there are three sites on the north east flank and
within each site there are five permanent transects. Every year, the
team return and take photos every meter along the transects. Once back
in the laboratory, five points from every photo are scored according
to what is represented underneath the point.

The main objective of long-term monitoring is to be able to report on
status and trends. Specifically, what is the status of each major
benthic group (such as hard coral, soft coral and macroalgae) and how
are they changing (particularly in response to major disturbances).

For this example, we will focus on a single reef (Agincourt Reef
No.1).

# Read in the data

```{r readData, results='markdown', eval=TRUE}
ltmp <- read_csv('../data/ltmp.csv', trim_ws=TRUE)
glimpse(ltmp)
```

| AIMS_REEF_NAME | REPORT_YEAR | SITE_NO | TRANSECT_NO | HC   | n.points | total.points |
|----------------|-------------|---------|-------------|------|----------|--------------|
| Arlington Reef | 2006        | 1       | 1           | 10.0 | 20       | 200          |
| Arlington Reef | 2006        | 1       | 2           | 10.5 | 21       | 200          |
|                |             |         |             |      |          |              |


# Data preparation


# Exploratory Data Analysis



# Fit the model

::: {.panel-tabset}


:::

# MCMC sampling diagnostics

::: {.panel-tabset}


:::

# Model validation

::: {.panel-tabset}


:::

# Model investigation

::: {.panel-tabset}


:::


# Further mnvestigations

::: {.panel-tabset}

:::


# As a GAM


```{r}
#| label: as a gam
#| results: markup
#| eval: false
#| echo: true
#| cache: false



ltmp_form <- bf(
    n.points | trials(total.points) ~ gp(REPORT_YEAR) +
        (1 | REEF_SITE) + (1 | REEF_SITE_TRANSECT),
    family = binomial(link = "logit")
)
get_prior(ltmp_form, data = ltmp_sub)
## we are not putting a prior on wiggliness directly.  Rather we are putting a prior on
## GP hyperparameters.  The GP hyperparameters are the length scale and the standard deviation of the GP.  The length scale controls how quickly the GP can change (smaller values allow for more rapid changes), and the standard deviation controls how much variation there is in the GP.  We will put a prior on these hyperparameters.
#- lscale: the length scale of the GP. Distance over which the GP function is correlated.  Smaller values allow for more rapid changes in the GP.  Larger values allow for smoother changes in the GP.
#- sdgp: how much can the trend deviate from the intercept (on link scale)

ltmp_sub |>
    summarise(
        Median = median(qlogis(HC/100)),
        MAD = mad(qlogis(HC/100)),
        N = mean(total.points)
    )
priors <- prior(normal(-1.1, 0.6), class = "Intercept") +
  prior(student_t(3, 0, 0.6), class = "sd") +
  prior(lognormal(log(5), 1), class = "lscale", coef = "gpREPORT_YEAR") +
  prior(exponential(5), class = "lscale", coef = "gpREPORT_YEAR") +
  prior(student_t(3, 0, 2), class = "sdgp")
  ## prior(gamma(2, 0.1), class = "lscale")
  ## prior(student_t(3, 0, 1/30), class = "lscale")
  ## prior(lognormal(log(5), 0.7), class = "lscale", coef = "gpREPORT_YEAR")
  ## prior(student_t(3, 0, 1), class = "lscale", coef = "gpREPORT_YEAR")
ltmp_sub_brm5 <- brm(ltmp_form,
                  data = ltmp_sub,
                  prior = priors,
                  sample_prior = 'yes',
                  iter = 3000,
                  warmup = 1000,
                  chains = 3, cores = 3,
                  thin = 10,
                  refresh = 100,
                  seed = 123,
                  control = list(adapt_delta = 0.99, max_treedepth = 20),
                  backend = "cmdstanr"
                  )

ltmp_sub_brm5 |>
  conditional_effects(conditions = data.frame(total.points = 200)) |>
  plot(points = TRUE)


ltmp_sub_brm5 |> summary()

newdata <- data.frame(
  REPORT_YEAR = seq(1995, 2023, len = 100),
  REEF_SITE = NA,
  REEF_SITE_TRANSECT = NA,
  total.points = 1
)
ltmp_sub_sum <-
  ltmp_sub_brm5 |>
    epred_draws(newdata = newdata, re_formula = NA) |>
  ungroup() |>
  dplyr::select(REPORT_YEAR, .epred) |>
  group_by(REPORT_YEAR) |>
  summarise_draws(median, HDInterval::hdi)

ltmp_sub_sum |>
  ggplot(aes(y = median, x = REPORT_YEAR)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "orange", alpha = 0.3) +
  geom_line() +
  geom_point(data = ltmp_sub,
    aes(y = HC/100))


ltmp_sub_brm5 |>
  epred_draws(newdata = newdata, re_formula = NA) |>
  ungroup() |>
  dplyr::select(.draw, REPORT_YEAR, .epred) |>
  ggplot(aes(y = .epred, x = REPORT_YEAR)) +
  geom_line(aes(group = .draw), alpha = 0.1, colour = "orange") +
  geom_line(data = ltmp_gp, inherit.aes = FALSE,
    aes(y = median, x = REPORT_YEAR), colour = "white", linewidth = 2) +
  geom_point(data = ltmp_sub,
    aes(y = HC/100, x = REPORT_YEAR))


forcast_newdata <- data.frame(
  REPORT_YEAR = seq(2023, 2030, len = 100),
  REEF_SITE = NA,
  REEF_SITE_TRANSECT = NA,
  total.points = 1
)

ltmp_sub_forcast_sum <-
  ltmp_sub_brm5 |>
    epred_draws(newdata = forcast_newdata, re_formula = NA) |>
  ungroup() |>
  dplyr::select(REPORT_YEAR, .epred) |>
  group_by(REPORT_YEAR) |>
  summarise_draws(median, HDInterval::hdi)

ltmp_sub_sum |>
    ggplot(aes(y = median, x = REPORT_YEAR)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "orange", alpha = 0.3) +
    geom_line() +
    geom_point(
        data = ltmp_sub,
        aes(y = HC / 100)
    ) +
  geom_ribbon(data = ltmp_sub_forcast_sum,
    aes(ymin = lower, ymax = upper), fill = "red", alpha = 0.3) +
  geom_line(data = ltmp_sub_forcast_sum)



forcast <- ltmp_sub_brm5 |>
    epred_draws(newdata = forcast_newdata, re_formula = NA) |>
    ungroup() |>
  dplyr::select(.draw, REPORT_YEAR, .epred)

ltmp_sub_brm5 |>
    epred_draws(newdata = newdata, re_formula = NA) |>
    ungroup() |>
    dplyr::select(.draw, REPORT_YEAR, .epred) |>
    ggplot(aes(y = .epred, x = REPORT_YEAR)) +
    geom_line(aes(group = .draw), alpha = 0.1, colour = "orange") +
    geom_line(
        data = ltmp_gp, inherit.aes = FALSE,
        aes(y = median, x = REPORT_YEAR), colour = "white", linewidth = 2
    ) +
    geom_point(
        data = ltmp_sub,
        aes(y = HC / 100, x = REPORT_YEAR)
    ) +
  geom_line(data = forcast,
    aes(group = .draw), alpha = 0.1, colour = "red")


forcast_sum <- forcast |>
  group_by(REPORT_YEAR) |>
  summarise_draws(median, HDInterval::hdi)

ltmp_sub_brm5 |>
    epred_draws(newdata = newdata, re_formula = NA) |>
    ungroup() |>
    dplyr::select(.draw, REPORT_YEAR, .epred) |>
    ggplot(aes(y = .epred, x = REPORT_YEAR)) +
    geom_line(aes(group = .draw), alpha = 0.1, colour = "orange") +
    geom_line(
        data = ltmp_gp, inherit.aes = FALSE,
        aes(y = median, x = REPORT_YEAR), colour = "white", linewidth = 2
    ) +
    geom_point(
        data = ltmp_sub,
        aes(y = HC / 100, x = REPORT_YEAR)
    ) +
  geom_line(data = forcast,
    aes(group = .draw), alpha = 0.1, colour = "red") +
  geom_line(
    data = forcast_sum, inherit.aes = FALSE,
    aes(y = median, x = REPORT_YEAR), colour = "grey", linewidth = 2
  )





newdata <- data.frame(
  REPORT_YEAR = c(1995, 2005),
  REEF_SITE = NA,
  REEF_SITE_TRANSECT = NA,
  total.points = 1
)
ltmp_sub_sum <-
  ltmp_sub_brm5 |>
  epred_draws(newdata = newdata, re_formula = NA) |>
  tidybayes::compare_levels(
    var = .epred, by = REPORT_YEAR,
    comparison = emmeans_comparison("revpairwise")
  ) |>
  ungroup() |>
  dplyr::select(REPORT_YEAR, .epred) |>
  group_by(REPORT_YEAR) |>
  summarise_draws(median, HDInterval::hdi)


  ltmp_sub_brm5 |>
  epred_draws(newdata = newdata, re_formula = NA) |>
  mutate(.epred = log(.epred)) |>
  tidybayes::compare_levels(
    var = .epred, by = REPORT_YEAR,
    comparison = emmeans_comparison("revpairwise")
  ) |>
  ungroup() |>
  dplyr::select(REPORT_YEAR, .epred) |>
  mutate(.epred = exp(.epred)) |>
  group_by(REPORT_YEAR) |>
  summarise_draws(median, HDInterval::hdi)

```
